Wrattler: \textnormal{A platform for AI-assisted data science}

### 2018-03-12 report/paper.tex
fixing typos

\paragraph{The double Anna Karenina principle.}
Not only is every dirty data set  dirty in its own way, but \emph{pace} Tolstoy, every clean data set is clean in its own way as well.
"Data" is such an abstract concept that specific integrity conditions to characterize whether data is dirty, and potentially even the most appropriate formalism for integrity conditions,  differ dramatically across
the vast array of disparate use cases of data science,
ranging form relational data describing the customers of a country, time series data describing sensor data in an internet of things platform, and huge datasets of satellite imagery of the earth over a multi-year time scale.

@@

\paragraph{Interactive.} Interactivity
is a necessary because ``big data is big''. The analyst learns about
the data set and the problem as she explores it.
Wrattler enables an efficient interaction by bringing computation closer to the human.
Notebooks run in the browser, cache partial results of computations and provide previews
of script results on-the-fly.

@@

Notebook systems such as Jupyter became a popular programming environment for data science, because
they support gradual data exploration and provide a convenient way of interleaving code with
comments and visualizations. However, notebooks suffer from a number of issues that hanper
reproducibility and limit the possible interaction model.

@@

The data store provides a way for persistently stroing data and enables communication between
individual Wrattler components. Data store keeps external data files imported into Wrattler,
as well as computed data frames (and, potentially, other data structures). It is immutable
and keeps past versions of data to allow efficient state rollback. For persistency and versioning,
data store also serializes and stores multiple versions of the dependency graph.

The data store supports a mechanism for annotating data frames with additional semantic information, such as:
%
\begin{itemize}
\item[--] Columns can be annotated with (and stored as) primitive data type such as date or floating-point number.
\vspace{0.25em}
\item[--] Columns (or a combination) can be annotated with semantic annotation such as geo-location or and address.
\vspace{-0.85em}
\item[--] Columns, rows and cells of the data frame can be annotated with other metadata such as provenance.
\end{itemize}

@@

Language runtimes are responsible for evaluating code and assisting with code analysis during
the construction of dependency graph. They can run as server-side components (e.g.~for R and Python)
or as client-side components (in case of JavaScript).
Unlike Jupyter kernels, a language runtime is stateless. It reads data from and writes data to
the data store. (Using a cache for efficiency.)

@@

Wrattler includes a scripting language with simple syntax that is used by individual AI assistants
to define domain specific languages. The syntax is shared across multiple AI assistnts, but the
operations are provided by individual assistants.

@@

The code produced by an AI assistant can be used in two ways. First, the analyst writes the code
interactively during an interaction with the AI assistant. During this phase, Wrattler
keeps a fine-grained representation of the code in the dependency graph (second cell in
Figure~\ref{fig:deps}), which allows Wrattler to provie live previews of the result, giving the
user an immediate feedback.

@@

\paragraph{Data integration} is a problem as common
as it is difficult. We envisage assistants
that combine logical and statistical reasoning to assist
with data  integration. A simple but useful
example of a tool in this space is to
 identifying pairs of variables across multiple data
sources that are likely to be daily temperatures.
More sophisticated examples could include and extend the
most cutting edge techniques in the data fusion and
semantic technologies literatures.

@@

\paragraph{Data cleaning} is a well studied
problem in the databases and data mining literature
\cite{abedjan2016detecting,ilyas2015},
but there are still a huge number of issues that have been unexplored.
For example, we envisage assistants for
Identifying values that seem to be out of range, e.g. missing value indicators;
Identifying possible inconsistent coding of strings,
such as different representations of acrynoms;
indentifying rows that seem "jointly" unusual, i.e., that seem not to conform to statistical dependencies among the columns; and
identifying distinct records that may refer to the same entity (record linkage).

\paragraph{Data monitoring.} Once an analysis
is completed, it may result in a machine learning
model which will then be deployed on a data stream.
Once a classifier is deployed, it must be monitored
to detect shifts in its input distribution,
such as change points, covariate shift, and
concept drift. This includes the problem
to suggesting when/how often new data points should be labelled for monitoring.
Even for data analyses that do not involve
deploying a ``production model'', there is
often a need to run an analysis periodically on updated data, such as re-running last quarter's
analysis on this quarter's data.
In such cases, we might believe that the two data
sets come from approximately the same distribution
and format, so we can exploit
divergences from this assumption to indicate
potential issues with data quality.
As part of the Wrattler effort, we have
introduced the \emph{data diff} problem \cite{datadiff},
which aims to improve the process of data wrangling by
returning a report of the differences between two data sets.

@@

\paragraph{Debugging and interpreting predictive analytics.}
Every time a new machine learning model is trained,
the first question an analyst has is often
why its accuracy is not better. But
debugging predictive analytics is notoriously difficult, and the remedies indirect and expensive,
 e.g., labelling more data, adding more features,
 using a more complex model. Debugging and improving
 models is a rich area for potential assistants.
 There is potentially low hanging fruit available
 in how to aid the tasks of
error analysis, exploring the
predictions of the model
on validation data. Some early work in this direction is
\cite{saleema}.
There is a rich and growing literature on
interpreting and explaining the predictions
of a model \cite{lipton:mythos,doshi-velez17,ribiero2016lime,darksight}.
Two potential directions from data are
debugging by using
"training set blame": Given an error in the validation data, what examples from the training set caused me to make that error? This might build on the work of \cite{percy}.
Another potential avenue is explaining
differences between predictions: why was instance 42 treated differently than instance 24601, even though they appear highly similar to a human?
}

@@

In this section, we focus on the initial functionality that is required from the data store.
We focus on the public interface exposed by the data store, but We do not discuss the technolgy
used for data storage -- this can be decided during the development.

Data frames stored in the data store can be annotated with metadata. Such metadata can
communicate information between multiple AI assistants. For example, first AI assistant can
infer that a type of a pair of columns represents GPS coordinate and produce a data frame
with annotations specifying this information. A second AI assistant can then use this information
and automatically provide a data visualization in the form of map. In the rest of this section,
we discuss the structure of those annotations.

@@

\item[--] \textbf{Geo coordinates}. Finally, the list of primitive types might also include geographical
  coordinates, in which case the data store will recognize basic formats such as GPS coordinates
  and Northings and Eastings.  
\end{itemize}

@@

In addition, Wrattler should define a set of annotations that can be used by a wide range of
standard data sets. It is useful to provide a basic schema for such common annotations as this
will enable a range of AI assistants and other tools to exchange information without first having
to define a common custom schema. Those should include annotations for places (addresses, 
countries, cities, postcodes), physical units (meters, killograms, etc.), contact information
(e-mails, phone numbers) and similar. We will investigate whether Schema.org or other existing
standards developed by the semantic web community already cover those areas.

>>>>

\paragraph{The double Anna Karenina principle.}
Not only is every dirty data set  dirty in its own way, but \emph{pace} Tolstoy, every clean data set is clean in its own way as well.
"Data" is such an abstract concept that specific integrity conditions to characterize whether data is dirty, and potentially even the most appropriate formalism for integrity conditions,  differ dramatically across
the vast array of disparate use cases of data science,
ranging from relational data describing the customers of a country, time series data describing sensor data in an internet of things platform, and huge datasets of satellite imagery of the earth over a multi-year time scale.

@@

\paragraph{Interactive.} Interactivity
is necessary because ``big data is big''. The analyst learns about
the data set and the problem as she explores it.
Wrattler enables an efficient interaction by bringing computation closer to the human.
Wrattler notebooks run in the browser, cache partial results of computations and provide previews
of script results on-the-fly.

@@

Notebook systems such as Jupyter have become a popular programming environment for data science, because
they support gradual data exploration and provide a convenient way of interleaving code with
comments and visualizations. However, notebooks suffer from a number of issues that hamper
reproducibility and limit the possible interaction model.

@@

The data store provides a way for persistently storing data and enables communication between
individual Wrattler components. Data store keeps external data files imported into Wrattler,
as well as computed data frames (and, potentially, other data structures). It is immutable
and keeps past versions of data to allow efficient state rollback. For persistency and versioning,
data store also serializes and stores multiple versions of the dependency graph.

The data store supports a mechanism for annotating data frames with additional semantic information, such as:
%
\begin{itemize}
\item[--] Columns can be annotated with (and stored as) primitive data type such as date or floating-point number.
\vspace{0.25em}
\item[--] Columns (or a combination) can be annotated with semantic annotation such as geo-location or address.
\vspace{-0.85em}
\item[--] Columns, rows and cells of the data frame can be annotated with other metadata such as provenance.
\end{itemize}

@@

Language runtimes are responsible for evaluating code and assisting with code analysis during
the construction of dependency graph. They can run as server-side components (e.g.~for R and Python)
or as client-side components (in case of JavaScript).
Unlike Jupyter kernels, a language runtime is stateless. It reads data from and writes data to
the data store (using a cache for efficiency).

@@

Wrattler includes a scripting language with simple syntax that is used
by individual AI assistants to define domain specific languages. The
syntax is shared across multiple AI assistants, but the operations are
provided by individual assistants.

@@

The code produced by an AI assistant can be used in two ways. First, the analyst writes the code
interactively during an interaction with the AI assistant. During this phase, Wrattler
keeps a fine-grained representation of the code in the dependency graph (second cell in
Figure~\ref{fig:deps}), which allows Wrattler to provide live previews of the result, giving the
user an immediate feedback.

@@

\paragraph{Data integration} is a problem as common
as it is difficult. We envisage assistants
that combine logical and statistical reasoning to assist
with data  integration. A simple but useful
example of a tool in this space is to
 identify pairs of variables across multiple data
sources that are likely to be daily temperatures.
More sophisticated examples could include and extend the
most cutting edge techniques in the data fusion and
semantic technologies literatures.

@@

\paragraph{Data cleaning} is a well studied
problem in the databases and data mining literature
\cite{abedjan2016detecting,ilyas2015},
but there are still a huge number of issues that have been unexplored.
For example, we envisage assistants for:
identifying values that seem to be out of range, e.g. missing value indicators;
identifying possible inconsistent coding of strings,
such as different representations of acrynoms;
identifying rows that seem "jointly" unusual, i.e., that seem not to conform to statistical dependencies among the columns; and
identifying distinct records that may refer to the same entity (record linkage).

\paragraph{Data monitoring.} Once an analysis
is completed, it may result in a machine learning
model which will then be deployed on a data stream.
Once a classifier is deployed, it must be monitored
to detect shifts in its input distribution,
such as change points, covariate shift, and
concept drift. This includes the problem
of suggesting when/how often new data points should be labelled for monitoring.
Even for data analyses that do not involve
deploying a ``production model'', there is
often a need to run an analysis periodically on updated data, such as
re-running last quarter's analysis on this quarter's data.
In such cases, we might believe that the two data
sets come from approximately the same distribution
and format, so we can exploit
divergences from this assumption to indicate
potential issues with data quality.
As part of the Wrattler effort, we have
introduced the \emph{data diff} problem \cite{datadiff},
which aims to improve the process of data wrangling by
returning a report of the differences between two data sets.

@@

\paragraph{Debugging and interpreting predictive analytics.}
Every time a new machine learning model is trained, the first question
an analyst has is often why its accuracy is not better. But debugging
predictive analytics is notoriously difficult, and the remedies
indirect and expensive, e.g., labelling more data, adding more
features, using a more complex model. Debugging and improving models
is a rich area for potential assistants.  There is potentially low
hanging fruit available in how to aid the tasks of error analysis,
exploring the predictions of the model on validation data. Some early
work in this direction is \cite{saleema}.  There is a rich and growing
literature on interpreting and explaining the predictions of a model
\cite{lipton:mythos,doshi-velez17,ribiero2016lime,darksight}.  Two
potential directions from are debugging by using "training set
blame": Given an error in the validation data, what examples from the
training set caused me to make that error? This might build on the
work of \cite{percy}.  Another potential avenue is explaining
differences between predictions: why was instance 42 treated
differently than instance 24601, even though they appear highly
similar to a human?  }

@@

In this section, we focus on the initial functionality that is
required from the data store.  We focus on the public interface
exposed by the data store, but we do not discuss the technology used
for data storage -- this can be decided during the development.

Data frames stored in the data store can be annotated with
metadata. Such metadata can communicate information between multiple
AI assistants. For example, first AI assistant can infer that a pair
of columns represent GPS lat/lon coordinates and produce a data frame
with annotations specifying this information. A second AI assistant
can then use this information and automatically provide a data
visualization in the form of map. In the rest of this section, we
discuss the structure of those annotations.

@@

\item[--] \textbf{Geo coordinates}. The list of primitive types might also include geographical
  coordinates, in which case the data store will recognize basic formats such as GPS coordinates  and Northings and Eastings.

\end{itemize}

@@

In addition, Wrattler should define a set of annotations that can be used by a wide range of
standard data sets. It is useful to provide a basic schema for such common annotations as this
will enable a range of AI assistants and other tools to exchange information without first having
to define a common custom schema. Those should include annotations for places (addresses, 
countries, cities, postcodes), physical units (meters, kilograms, etc.), contact information
(e-mails, phone numbers) and similar. We will investigate whether Schema.org or other existing
standards developed by the semantic web community already cover those areas.
